Non-convergence in iterative imputation

### 2025-05-23 Manuscript/manuscript.qmd
text flow

Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \gtrapprox 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \gtrapprox 0$). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold. The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13]. @vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$. The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15]. $AC$-values that are significantly higher than zero indicate non-stationarity.
<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->

@@

We evaluate the pooled estimates against their true population values using the performance measures bias, confidence interval width, and coverage rate, as recommended by van Buuren (-@buur18 $\S$ 2.5.2). We calculate bias as $\bar{Q} - Q$. CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$. CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

@@

Autocorrelation in the chain means decreases rapidly in the first few iterations (see @fig-ac_mean). The decrease is substantive until $n_{it} \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.

@@

## Conclusion

@@

## Limitations

@@

>>>>

Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \approx 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \approx 0$). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold. The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13]. @vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$. The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15]. $AC$-values that are significantly higher than zero indicate non-stationarity.
<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->

@@

We evaluate the pooled estimates against their true population values using the performance measures bias, confidence interval width, and coverage rate, as recommended by van Buuren (-@buur18 $\S$ 2.5.2). We calculate bias as $\bar{Q} - Q$. CR is defined as the proportion of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$. CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

@@

[TODO: edit!] Autocorrelation in the chain means decreases rapidly in the first few iterations (see @fig-ac_mean). The decrease is substantive until $n_{it} \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.

@@

@@

[TODO: add segway]

@@

# References
